### 2015-05-15 corral.tex
fixes

\whoop is a sound but imprecise static data race analyzer. For developers who deem false alarms as unacceptable, we consider a method for leveraging the full or partial race-freedom guarantees provided by \whoop to accelerate the performance of \corral~\cite{lal2012corral}, a precise bug-finder. \corral, which is used by Microsoft to analyze Windows drivers~\cite{lal2014powering}, has capabilities for analyzing concurrent programs.  Because \corral operates on Boogie programs, it was easy to integrate it into our toolchain (see Figure~\ref{fig:whoop} -- C). Our technique, though, is general and capable in principle of accelerating any concurrency bug-finder that operates by interleaving threads at shared memory operations.

>>>>

\whoop is a sound but imprecise static data race analyzer. For developers who deem false alarms as unacceptable, we consider a method for leveraging the full or partial race-freedom guarantees provided by \whoop to accelerate the performance of \corral~\cite{lal2012corral}, a precise bug-finder used by Microsoft to analyze Windows drivers~\cite{lal2014powering}.  Because \corral operates on Boogie programs, it was easy to integrate it into our toolchain (see Figure~\ref{fig:whoop} -- C). Our technique, though, is general and capable in principle of accelerating any concurrency bug-finder that operates by interleaving threads at shared memory operations.

### 2015-05-15 evaluation.tex
fixes

We performed experiments to validate the usefulness of the \whoop approach (\S\ref{whoop}) and its combination with \corral (\S\ref{corral}). We first present race-checking results from running \whoop and \corral on \sizeOfBenchmarks drivers taken from the Linux 4.0 kernel.\footnote{\url{https://www.kernel.org}} We then evaluate the runtime performance and scalability of \corral and \whoop + \corral with different yield instrumentation granularities and context-switch bounds.
Our results demonstrate that \whoop can efficiently accelerate race-checking with \corral.

>>>>

We performed experiments to validate the usefulness of the \whoop approach (\S\ref{whoop}) and its combination with \corral (\S\ref{corral}). We first present race-checking results from running \whoop and \corral on \sizeOfBenchmarks drivers taken from the Linux 4.0 kernel.\footnote{\url{https://www.kernel.org}} We then evaluate the runtime performance and scalability of \corral using different yield instrumentations and context-switch bounds. Our results demonstrate that \whoop can efficiently accelerate race-checking with \corral.

### 2015-05-15 whoop.tex
fixes

We now present \emph{symbolic pairwise lockset analysis}, a novel method for data race analysis in device drivers.  In \S\ref{sec:symbolicpairwise} we describe how the approach works in a semi-formal manner, with respect to a simple concurrent programming model.  In \S\ref{sec:implementation} we explain how we have implemented symbolic pairwise lockset analysis in a practical tool, \whoop, that can be applied directly to driver source code.
%
Our experimental evaluation (\S\ref{evaluation}) demonstrates that the \whoop technique has value as a stand-alone analyzer, and that results from \whoop can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency, based on context bounding and stratified inlining, offered by the \corral~\cite{lal2012corral} tool; we discuss the latter approach in \S\ref{corral}.

@@

The forms of statements are shown in Figure~\ref{fig:statements}, and include designated statements for reading from and writing to shared variables.  In particular, shared variables may not appear in arbitrary expressions.  This restriction simplifies our presentation of lockset instrumentation below, and a program that does not satisfy this restriction can be trivially pre-processed into one that does via the introduction of additional private temporary variables to record values read from the shared state.  We do not specify the form of expressions, nor the types of variables, assuming a standard set of data types and operations.

\begin{figure}
\footnotesize
\center
\begin{tabular}{lp{5.5cm}}
\textbf{Statement Form} & \textbf{Notes} \\
\toprule

@@

$\mutexunlock{m};$ & mutex unlock, where $m \in \mutexes$\\
\bottomrule
\end{tabular}
\caption{Forms of statements in our simple programming model}
\label{fig:statements}
\vspace{-2mm}
\end{figure}

@@

At the start of execution the valuation of shared variables is arbitrary, no mutexes are held (i.e.\ each mutex maps to $\bot$), and the list of threads is empty.
%
At any point of execution, a new thread may be added to the list of threads.  This involves selecting a thread template $T$ and an id $i \in \ids$ that has not been previously used during program execution, setting the point of execution for the new thread to be the first statement of $\mainforthread{T}$, and choosing an arbitrary valuation for the private variables $\variables{T}$.
%
We consider a standard interleaving model of concurrency: at any execution point, a thread may execute its current statement, unless that statement has the form $\mutexlock{m}$ and mutex $m$ is already held by some thread.  Executing a statement causes the state of the thread, and the shared state, to be updated in a standard manner.  For example, if a thread following template $T$ executes $s = e$, where $s \in \sharedvars$ and $e$ is an expression over $\variables{T}$, the shared variable valuation is updated so that $s$ has the value determined by evaluating $e$ in the context of the thread's private variable valuation.  Because our interest is in data race analysis for race-free programming, we are not concerned with relaxed memory behavior: race-free programs exhibit only sequentially consistent behaviors.

@@

$\mutexunlock{m};$ & $\currentlockset{i} = \currentlockset{i} \setminus \{ m \};$ \\
\bottomrule
\end{tabular}
\caption{Instrumenting statements for lockset analysis}
\label{fig:instrumentation}
\end{figure}

@@

\noindent\textbf{Soundness }
%
We sketch an argument that if the program of Figure~\ref{fig:sequentialization} is correct, it is impossible for a thread executing template $T$ to race with a thread executing template $U$, under the assumption that the threads are guaranteed to terminate. Let us assume that the program of Figure~\ref{fig:sequentialization} is correct, and suppose (by way of contradiction) that a thread executing $T$ can in fact race with a thread executing $U$, on some shared variable $s$.  By our hypothesis that the program of Figure~\ref{fig:sequentialization} is correct, and that the threads terminate, the assertion checked at the end of the program guarantees at least one mutex, say $m$, belongs to both $\lockset{s}{i}$ and $\lockset{s}{j}$.  By the definition of a lockset (and according to the manner in which shared accesses are instrumented in Figure~\ref{fig:instrumentation}), this means that $m$ is held during every access to $s$ by both $\instrument{T}{i}$ and $\instrument{U}{j}$. As a result, $m$ must be unlocked and locked between the two accesses, which contradicts that the pair of accesses is racing.

>>>>

We now present \emph{symbolic pairwise lockset analysis}, a novel method for data race analysis in device drivers.  In \S\ref{sec:symbolicpairwise} we describe how the approach works in a semi-formal manner, with respect to a simple concurrent programming model.  In \S\ref{sec:implementation} we explain how we have implemented our analysis in a practical tool, \whoop, that can be applied directly to driver source code.
%
Our experimental evaluation (\S\ref{evaluation}) demonstrates that \whoop has value as a stand-alone analyzer, and that results from our analysis can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency, offered by the \corral~\cite{lal2012corral} tool; we discuss the latter approach in \S\ref{corral}.

@@

The allowed statements are shown in Figure~\ref{fig:statements}, and include designated statements for reading from and writing to shared variables.  In particular, shared variables may not appear in arbitrary expressions.  This restriction simplifies our presentation of lockset instrumentation below, and a program that does not satisfy this restriction can be trivially pre-processed into one that does via the introduction of additional private temporary variables to record values read from the shared state.  We do not specify the form of expressions, nor the types of variables, assuming a standard set of data types and operations.

\begin{figure}
\footnotesize
\center
\setlength{\tabcolsep}{0.45em}
\begin{tabular}{lp{6.8cm}}
\textbf{Statement} & \textbf{Notes} \\
\toprule

@@

$\mutexunlock{m};$ & mutex unlock, where $m \in \mutexes$\\
\bottomrule
\end{tabular}
\caption{The allowed statements in our simple programming model}
\label{fig:statements}
\vspace{-2mm}
\end{figure}

@@

Initially, the valuation of shared variables is arbitrary, no mutexes are held (i.e.\ each mutex maps to $\bot$), and the list of threads is empty.
%
At any point of execution, a new thread may be added to the list of threads.  This involves selecting a thread template $T$ and an id $i \in \ids$ that has not been previously used during program execution, setting the point of execution for the new thread to be the first statement of $\mainforthread{T}$, and choosing an arbitrary valuation for the private variables $\variables{T}$.
%
We consider a standard interleaving model of concurrency: at any execution point, a thread may execute its current statement, unless that statement has the form $\mutexlock{m}$ and mutex $m$ is held by some thread.  Executing a statement causes the state of the thread, and the shared state, to be updated in a standard manner.  For example, if a thread following template $T$ executes $s = e$, where $s \in \sharedvars$ and $e$ is an expression over $\variables{T}$, the shared variable valuation is updated so that $s$ has the value determined by evaluating $e$ in the context of the thread's private variable valuation.  Because our interest is in data race analysis for race-free programming, we are not concerned with relaxed memory behavior: race-free programs exhibit only sequentially consistent behaviors.

@@

$\mutexunlock{m};$ & $\currentlockset{i} = \currentlockset{i} \setminus \{ m \};$ \\
\bottomrule
\end{tabular}
\caption{Instrumenting statements for lockset analysis}
\label{fig:instrumentation}
\vspace{-2mm}
\end{figure}

@@

\noindent\textbf{Soundness }
%
We sketch an argument that if the program of Figure~\ref{fig:sequentialization} is correct (i.e.\ the assertion, which is --- described above --- at the end of the program holds), then it is impossible for a thread executing template $T$ to race with a thread executing template $U$, under the assumption that the threads are guaranteed to terminate. Let us assume that the program of Figure~\ref{fig:sequentialization} is correct, and suppose (by way of contradiction) that a thread executing $T$ can in fact race with a thread executing $U$, on some shared variable $s$.  By our hypothesis that the program is correct, and that the threads terminate, the assertion checked at the end of the program guarantees that at least one mutex, say $m$, belongs to both $\lockset{s}{i}$ and $\lockset{s}{j}$.  By the definition of a lockset (and according to the manner in which shared accesses are instrumented in Figure~\ref{fig:instrumentation}), this means that $m$ is held during every access to $s$ by both $\instrument{T}{i}$ and $\instrument{U}{j}$. As a result, $m$ must be unlocked and locked between the two accesses, which contradicts that the pair of accesses is racing.
